The Unicode Cookbook \linebreak for Linguists

### 2018-03-06 book/chapters/introduction.tex
these changes almost complete the remaining reviews

Because 256 characters were clearly not enough for the many different characters
needed in the world's writing systems, the ISO/IEC~8859 standard in 1987
extended the MCS to include 16 different 8-bit code pages. For example, part 5
was used for Cyrillic characters, part 6 for Arabic, and part 7 for
Greek.\footnote{In effect, because $16 = 2^4$, this means that ISO/IEC~8859 was
actually an $8+4=12$-bit encoding, though with very many duplicates by design,
namely all ASCII codes were repeated in each 8-bit code page. To be precise,
ISO/IEC~8859 used the 7-bit ASCII as the basis for each code page, and defined
16 different 7-bit extensions, leading to $(1+16) \cdot {2^7} = 2,176$ possible
characters. However, because of overlap and not-assigned codes points the actual
number of symbols was much smaller.} This system almost immediately was
understood to be insufficient and impractical, so various initiatives to extend
and reorganize the encoding started in the 1980s. This led, for example, to
various proprietary encodings from Microsoft (e.g.~Windows Latin 1) and Apple
(e.g.~Mac OS Roman), which one still sometimes encounters today.

@@

Breaking it down further, a script consists of \textsc{graphemes}, and graphemes
consist of \textsc{characters}. In the linguistic terminology of writing
systems, a \textsc{character} is a general term for any self-contained element
in a writing system.\footnote{There is a second interpretation of the term
\textsc{character}, i.e.~a conventional term for a unit in the Chinese writing
system \citep{Daniels1996}. This interpretation will not be further explored in
this paper.} Although in literate societies most people have a strong intuition
about what the characters are in their particular orthography or orthographies,
it turns out that the separation of an orthography into separate characters is
far from trivial. The widespread intuitive notion of a character is strongly
biased towards educational traditions, like the alphabet taught at schools, and
technological possibilities, like the available type pieces in a printer's job
case, the keys on a typewriter, or the symbols displayed in Microsoft Word's
symbol browser. In practice, characters often consist of multiple building
blocks, each of which could be considered a character in its own right. For
example, although a Chinese character may be considered to be a single basic
unanalyzable unit, at a more fine-grained level of analysis the internal
structure of Chinese characters is often comprised of smaller semantic and
phonetic units that should be considered characters \citep{Sproat2000}. In
alphabetic scripts, this problem is most forcefully exemplified by diacritics.

A \textsc{diacritic} is a mark, or series of marks, that may be above, below, or
through other characters \citep{Gaultney2002}. Diacritics are sometimes used to
distinguish homophonous words, but they are more often used to indicate a
modified pronunciation \citep[xli]{DanielsBright1996}. The central question is
whether, for example, <e>, <è>, <a> and <à> should be considered four
characters, or different combinations of three characters. In general, multiple
characters together can form another character, and it is not always possible to
decide on principled grounds what should be the basic building blocks of an
orthography.

For that reason, it is better to analyze an orthography as a collection of
graphemes. A \textsc{grapheme} is the basic, minimally distinctive symbol of a
particular writing system, alike to the phoneme is an abstract representation of
a distinct sound in a specific language. The term \textsc{grapheme} was modeled
after the term \textsc{phoneme} and represents a contrastive graphical unit in a
writing system (see \citealp{Kohrt1986} for a historical overview of the term
grapheme). Most importantly, a single grapheme regularly consists of multiple
characters, like <th>, <ou> and <gh> in English (note that each character in
these graphemes is also a separate grapheme in English). Such complex graphemes
are often used to represent single phonemes. So, a combination of characters is
used to represent a single phoneme. Note that the opposite is also found in
writing systems, in cases in which a single character represents a combination
of two or more phonemes. For example, <x> in English orthography represents a
combination of the phonemes /k/ and /s/.

>>>>

Because 256 characters were clearly not enough for the many different characters
needed in the world's writing systems, the ISO/IEC~8859 standard in 1987
extended the MCS to include 16 different 8-bit code pages. For example, part 5
was used for Cyrillic characters, part 6 for Arabic, and part 7 for
Greek. % \footnote{In effect, because $16 = 2^4$, this means that ISO/IEC~8859 was
%actually an $8+4=12$-bit encoding, though with very many duplicates by design,
%namely all ASCII codes were repeated in each 8-bit code page. To be precise,
%ISO/IEC~8859 used the 7-bit ASCII as the basis for each code page, and defined
%16 different 7-bit extensions, leading to $(1+16) \cdot {2^7} = 2,176$ possible
%characters. However, because of overlap and not-assigned codes points the actual
%number of symbols was much smaller.} 
This system almost immediately was
understood to be insufficient and impractical, so various initiatives to extend
and reorganize the encoding started in the 1980s. This led, for example, to
various proprietary encodings from Microsoft (e.g.~Windows Latin 1) and Apple
(e.g.~Mac OS Roman), which one still sometimes encounters today.

@@

Breaking it down further, a script consists of \textsc{graphemes}, which are writing 
system-specific minimally distinctive symbols (see below). Graphemes may consist of one or more 
\textsc{characters}. The term \textsc{character} is overladen. In the linguistic terminology of writing
systems, a \textsc{character} is a general term for any self-contained element
in a writing system. A second interpretation is used as a conventional term for a unit in the Chinese writing
system \citep{Daniels1996}. In technological terminology, a \textsc{character} 
refers to the electronic encoding of a component in a writing system that has semantic 
value (see Section \ref{character-encoding-system}). Thus in this work we must navigate 
between the general linguistic and technological terms for \textsc{character} 
and \textsc{grapheme} because of how these notions are defined and how they relate at the intersection 
between the International Phonetic Alphabet and the Unicode Standard (Chapter \ref{ipa-meets-unicode}).

Although in literate societies most people have a strong intuition
about what the characters are in their particular orthography or orthographies,
it turns out that the separation of an orthography into separate characters is
far from trivial. The widespread intuitive notion of a character is strongly
biased towards educational traditions, like the alphabet taught at schools, and
technological possibilities, like the available type pieces in a printer's job
case, the keys on a typewriter, or the symbols displayed in Microsoft Word's
symbol browser. In practice, characters often consist of multiple building
blocks, each of which could be considered a character in its own right. For
example, although a Chinese character may be considered to be a single basic
unanalyzable unit, at a more fine-grained level of analysis the internal
structure of Chinese characters is often comprised of smaller semantic and
phonetic units that should be considered characters \citep{Sproat2000}. In
alphabetic scripts, this problem is most forcefully exemplified by diacritics.

A \textsc{diacritic} is a mark, or series of marks, that may be above, below, 
before, after, through, around, or between other characters \citep{Gaultney2002}. Diacritics are sometimes used to
distinguish homophonous words, but they are more often used to indicate a
modified pronunciation \citep[xli]{DanielsBright1996}. The central question is
whether, for example, <e>, <è>, <a> and <à> should be considered four
characters, or different combinations of three characters, i.e.\ <a>, <e>, and <\dia{0300}>. 
In general, multiple characters together can form another character, and it is not always possible to
decide on principled grounds what should be the basic building blocks of an
orthography.

For that reason, it is better to analyze an orthography as a collection of
graphemes. A \textsc{grapheme} is the basic, minimally distinctive symbol of a
particular writing system, alike to the phoneme is an abstract representation of
a distinct sound in a specific language. The term \textsc{grapheme} was modeled
after the term \textsc{phoneme} and represents a contrastive graphical unit in a
writing system (see \citealp{Kohrt1986} for a historical overview of the term
grapheme). Most importantly, a single grapheme regularly consists of multiple
characters, like <th>, <ou> and <gh> in English (note that each character in
these graphemes is also a separate grapheme in English). Such complex graphemes
are often used to represent single phonemes. So, a combination of characters is
used to represent a single phoneme. Note that the opposite is also found in
writing systems, in cases in which a single character represents a combination
of two or more phonemes. For example, <x> in English orthography sometimes represents a
combination of the phonemes /k/ and /s/, as in the word `index' [ˈɪnˌdɛks]; other times 
it is pronounced as /z/, such as in the words `Xerox' [ˈzɪrˌɑks] and `example' [ɪɡˈzæmpəl]. 
As one can see, there can be non-trivial mappings between graphemes and phonemes in 
orthographies like English, e.g.\ `give', `gin', `jingle', where the graphemes 
<g> and <j> and the phonemes /g/ and /dʒ/ have a complex mapping.

### 2018-03-06 book/chapters/ipa_background.tex
these changes almost complete the remaining reviews

Essentially transcription has two parts: a text containing symbols and a set 
of conventions for interpreting those symbols (and their combinations). 
The symbols of the IPA distinguish between letter-like symbols and
diacritics (symbol modifiers). The use of the letter-like symbols to represent 
a language's sounds is guided by the principle of contrast; where two words 
are distinguishable by phonemic contrast, those contrasts should be transcribed 
with different letter symbols (and not just diacritics). Allophonic distinction 
falls under the rubric of diacritically-distinguished symbols, e.g.\ the 
difference in English between an aspirated /p/ and /t/ in [spʰot̚] and 
an unrealeased /p/ in [stʰop̚].

@@

The interpretation of the IPA symbols in specific usage is not trivial. Although
the articulatory properties of the IPA symbols themselves are rather succinctly
summarised by the normative description in the \textit{Handbook}, it is common
in practical applications that the transcribed symbols do not precisely
represent the phonetic value of the sounds that they represent. So an IPA symbol
<t> in one transcription is not always the same as an IPA <t> in another
transcription (or even within a single transcription). The interpretation of any
particular <t> is mostly a language-specific convention (or even
resource-specific and possibly even context-specific), a fact which --
unfortunately -- is in most cases not made explicit by users of the IPA.

@@

In contrast, the basic goal of a phonemic broad transcription is to distinguish all
words in a language with a minimal number of transcription symbols
\citep[19]{Abercrombie1964}. A phonemic transcription includes the conventions
of a particular language's phonological rules. These rules determine the
realization of that language's phonemes. For example, in the transcription of
German, Dutch, English and French a symbol <t> might be used for the voiceless
plosive in the alveolar area. This symbol is sufficient for a succinct
transcription of these languages because there is no further phonemic
subdivisions in this domain in either of these languages. However, the
language-specific realization of this consonant is closer to [t̪ʰ], [t], [tʰ]
and [t̪], respectively. Similarly, the five vowels of Greek can be represented
phonemically in IPA as /ieaou/, though phonetically they are closer to [iεaɔu].
The Japanese five vowel system can also be transcribed in IPA as
/ieaou/, while the phonetic targets in this case are closer to [ieaoɯ].

@@

% \footnote{The marking of tone was extended (from characteristically Africanist practice) to include a second system for marking linguistic tones (the `Chao tone letters'), a much used convention based on musical staff to describe pitch in by Yuen Ren Chao (1892--1982).} 
\subsection*{IPA Numbers}
Prior to the Kiel Convention for the modern revision of the IPA in 1989,
\citet{Wells1987} collected and published practical approaches to coded
representations of the IPA, which dealt mainly with the assignment of characters
on the keyboard to IPA symbols. The process of assigning standardized computer
codes to phonetic symbols was given to the \textsc{workgroup on computer
coding} (henceforth \textit{working group}) at the Kiel Convention. This working
group was assigned to consider the following tasks
\citep{Esling1990,EslingGaylord1993}:

@@

The working group decided insightfully that the computing-coding convention 
for the IPA should be independent of computer environments or formats (e.g.\ ASCII), 
i.e.\ the IPA Number was not meant to be implemented directly in a computer 
encoding. The working group report's declaration includes the following explanatory 
remarks \citep[82]{International1989report}:

@@

Character set ISO/IEC 10646 was approved by ISO, including the phonetic
characters submitted to them in May 1993. The set of IPA characters were
assigned UCS codes in 16 bit representation (in hexadecimal) and were published
as Tables 2 and 3 in \cite{EslingGaylord1993}, which include a graphical
representation of the IPA Symbol, its IPA Name, phonetic description, IPA
Number, UCS Code and AFII Code.\footnote{The \textsc{association for font
information interchange} (AFII) was an international database of glyphs created
to promote the standardization of font data required to produce ISO/IEC 10646.} When the
character sets of ISO/IEC 10646 and the Unicode Standard later converged (see
Section~\ref{the-unicode-approach}), the IPA proposal was
included in the Unicode Standard Version 1.0 -- largely as we know it
today.

@@

It was a long journey, but the goal of achieving a single multilingual computing
environment has largely been accomplished. Current users, however, must cope
with the pitfalls that were dug along the way, as will be discussed in the next
chapter. As the Association foresightfully remarked about Unicode:

>>>>

Essentially transcription has two parts: a text containing symbols and a set 
of conventions for interpreting those symbols (and their combinations). 
The symbols of the IPA distinguish between letter-like symbols and
diacritics (symbol modifiers). The use of the letter-like symbols to represent 
a language's sounds is guided by the principle of contrast; where two words 
are distinguishable by phonemic contrast, those contrasts should be transcribed 
with different letter symbols (and not just diacritics). Allophonic distinction 
falls under the rubric of diacritically-distinguished symbols, e.g.\ the 
difference in English between an aspirated /p/ and /t/ in [spʰot̚] and 
an unreleased /p/ in [stʰop̚].

@@

The interpretation of the IPA symbols in specific usage is not trivial. Although
the articulatory properties of the IPA symbols themselves are rather succinctly
summarized by the normative description in the \textit{Handbook}, it is common
in practical applications that the transcribed symbols do not precisely
represent the phonetic value of the sounds that they represent. So an IPA symbol
<t> in one transcription is not always the same as an IPA <t> in another
transcription (or even within a single transcription). The interpretation of any
particular <t> is mostly a language-specific convention (or even
resource-specific and possibly even context-specific), a fact which --
unfortunately -- is in most cases not made explicit by users of the IPA.

@@

In contrast, the basic goal of a phonemic broad transcription is to distinguish all
words in a language with a minimal number of transcription symbols
\citep[19]{Abercrombie1964}. A phonemic transcription includes the conventions
of a particular language's phonological rules. These rules determine the
realization of that language's phonemes. For example, in the transcription of
German, Dutch, English and French a symbol <t> might be used for the voiceless
plosive in the alveolar and dental areas. This symbol is sufficient for a succinct
transcription of these languages because there is no further phonemic
subdivisions in this domain in either of these languages. However, the
language-specific realization of this consonant is closer to [t̪ʰ], [t], [tʰ]
and [t̪], respectively. Similarly, the five vowels of Greek can be represented
phonemically in IPA as /ieaou/, though phonetically they are closer to [iεaɔu].
The Japanese five vowel system can also be transcribed in IPA as
/ieaou/, while the phonetic targets in this case are closer to [ieaoɯ].

@@

% \footnote{The marking of tone was extended (from characteristically Africanist practice) to include a second system for marking linguistic tones (the `Chao tone letters'), a much used convention based on musical staff to describe pitch in by Yuen Ren Chao (1892--1982).}

\subsection*{IPA Numbers}
Prior to the Kiel Convention for the modern revision of the IPA in 1989,
\citet{Wells1987} collected and published practical approaches to coded
representations of the IPA, which dealt mainly with the assignment of characters
on the keyboard to IPA symbols. The process of assigning standardized computer
codes to phonetic symbols was given to the \textsc{workgroup on computer
coding} (henceforth \textit{working group}) at the Kiel Convention. This working
group was assigned to consider the following tasks
\citep{Esling1990,EslingGaylord1993}:

@@

The working group decided insightfully that the computing-coding convention 
for the IPA should be independent of computer environments or formats, 
i.e.\ the IPA Number was not meant to be encoded at the bit pattern level.
The working group report's declaration includes the following explanatory 
remarks \citep[82]{International1989report}:

@@

Character set ISO/IEC 10646 was approved by ISO, including the phonetic
characters submitted to them in May 1993. The set of IPA characters were
assigned UCS codes in 16 bit representation (in hexadecimal) and were published
as Tables 2 and 3 in \cite{EslingGaylord1993}, which include a graphical
representation of the IPA Symbol, its IPA Name, phonetic description, IPA
Number, UCS Code and AFII Code.\footnote{The \textsc{association for font
information interchange} (AFII) was an international database of glyphs created
to promote the standardization of font data required to produce ISO/IEC 10646.} When the
character sets of ISO/IEC 10646 and the Unicode Standard later converged (see
Chapter~\ref{the-unicode-approach}), the IPA proposal was
included in the Unicode Standard Version 1.0 -- largely as we know it
today.

@@

It was a long journey, but the goal of achieving a single multilingual computing
environment has largely been accomplished. As such, we should not dismiss the IPA Numbers 
or pre-Unicode encoding attempts, such as SAMPA/X-SAMPA, as misguided. The parallels 
between the IPA Numbers and Unicode Code points, for example, is striking because both the IPA and 
the Unicode Consortium came up with the solution of an additional layer of indirection (an abstraction layer) 
between symbols/characters and encoding on the bit pattern level. SAMPA/X-SAMPA is also still useful 
as an input method for IPA in ASCII and required by some software.

Current users of the Unicode Standard must cope
with the pitfalls that were dug along the way, as will be discussed in the next
chapter. As the Association foresightfully remarked about Unicode:
